---
title: "NICE visual cookbook for R graphics"
author: "Impact team"
date: "Last updated: November 2022"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dev = "ragg_png")
here::i_am("reports/NICE_cookbook.Rmd")

```


# How to create NICE style graphics

The `niceRplots` package and cookbook make the process of creating publication-ready graphics in the NICE style a more reproducible process, as well as making it easier for people new to R to create graphics. 

Within this cookbook, we will demonstrate how to use the functions included within the `niceRplots` package. We will also provide examples of several plots that have been created in the NICE style. 

## Load the relevant packages

Before we can start making charts, we first need to load all of the relevant packages. We wont load every package here, just the ones that we will use the most often. It is good practice to limit the number of loaded packages to prevent masking conflicts. For this reason, we will often use double colons `::` to access the specific functions within a package (e.g. `readr::read_csv()`), rather than loading the full package using `library(pkg)`.

```{r pkg_load, message = FALSE}

# Load in the relevant packages
library(dplyr)
library(ggplot2)
library(plotly)
library(leaflet)

devtools::load_all()

# Set up the NICE colour palette
nice_colours <- c("#228096",
                  "#00436C",
                  "#37906D",
                  "#D07B4D",
                  "#EAD054")

# nice_colours <- c("Bold teal" = "#228096",
#                   "Deep blue" = "#00436C",
#                   "Balanced green" = "#37906D",
#                   "Natural tan" = "D07B4D",
#                   "Positive yellow" = "EAD054")

```


## How to style your charts

As an example we will create a simple bar chart using the iris dataset. We will first create a basic chart, and then add some additional formatting. We can see the two charts below with and without formatting.

```{r format_chart, echo = TRUE, fig.show="hold", out.width="50%"}

# Wrangle data for plotting
iris_bar_df <- iris %>% 
  group_by(Species) %>% 
  summarise(Sepal_Width = mean(Sepal.Width)) %>% 
  ungroup()

# Create basic bar chart
iris_bar <- ggplot(iris_bar_df) + 
  geom_col(aes(x = Species, y = Sepal_Width), ) +
  labs(title = "Basic chart title",
       subtitle = "Basic chart subtitle",
       x = "Species",
       y = "Sepal Width")

# Create formatted chart
iris_bar_formatted <- ggplot(iris_bar_df) + 
  # Adjust bar fill and outline colour
  geom_col(aes(x = Species, y = Sepal_Width), fill = "#228096", colour = "#000000")+
  # Add baseline
  geom_hline(yintercept = 0, linewidth = 1, colour="#333333") +
  # set y-axis limits and remove padding
  scale_y_continuous(expand = c(0,0), limits = c(0,4)) +
  # Add labels 
  labs(title = "Formatted chart title",
       subtitle = "Formatted chart subtitle ",
       y = "Sepal width")

iris_bar
iris_bar_formatted

```


We can now apply the nice theme using the `nice_gg_simple()` function, specifying the base font for the plot to be of size 12. The `nice_gg_simple()` function alters several aspects of the chart to make it consistent with the [NICE brand guidelines](https://indepth.nice.org.uk/our-brand/index.html). These are as follows:

* **Fonts**: changes the title font to Lora SemiBold, and all other text to Inter Regular. 

* **Text scaling**: applies a text hierarchy to the chart using relative text sizes. Changing the `base_size` argument will change the text size throughout the plot. Titles and subtitles will be automatically scaled relative to this base_size to maintain the text hierarchy. This is useful if you want to change the chart size, as all text can be simply adjusted by changing the `base_size` argument. 

* **Colour scheme**: changes line and text colours to be consistent with the NICE brand colour palette. This is particularly noticeable when creating faceted charts (shown later in cookbook).

* **Margins**: adjusts the margins around the title, subtitle, and axis titles to give everything space to breathe.

* **Background**: removes the panel background and borders.

* **Other**: Applies major grid lines along the y-axis. This will be appropriate for most charts, but may need to be disabled in others by applying `theme(panel.grid.major.y = ggplot2::element_blank())`


The `nice_gg_theme` function isn't designed to modify every aspect of the chart, but to apply a general theme that will maintain consistency with the NICE style guide across all charts. Many aspects of the chart will still need to be modified manually. For example, in the below chart we may choose to apply an additional theme argument to remove the x-axis title. 

```{r theme_chart, echo = TRUE}

# Apply the NICE theme
iris_bar_themed <- iris_bar_formatted +
  # Apply nice theme, set base text size to 12
  nice_gg_simple(base_size = 12) +
  # remove x axis title as this is not needed
  theme(axis.title.x = ggplot2::element_blank())

iris_bar_themed

```


## How to finalise your charts

Once a plot has been created and styled using `NICE_gg_simple()`, the next step is to apply the `finalise_plot()` function. This function will create a footer containing information on the data source, as well as adding the NICE logo to the bottom right. It will then left-align the title, subtitle and footer. The `finalise_plot()` function has 4 arguments:

```{r finalise, echo = TRUE}

# Apply the finalise plot function
finalised_plot <- finalise_plot(iris_bar_themed,
              source_name = "Source: Iris dataset")

finalised_plot
```


## How to save your finalised plots

After applying the `finalise_plot()` function, the plot can be saved using the `ggsave()` function.


```{r save_plot, echo = TRUE, eval = FALSE}

# Save plot using the ggsave() function 
ggsave("finalised_plot.png", finalised_plot)

```


# Static charts using ggplot2

Before making charts we first have to read in our data. We will use data on monthly CCG level prescribing of 5 direct oral anticoagulants (DOACS) in England between January 2020 and December 2021. This was downloaded from [OpenPrescribing.net](https://openprescribing.net/).

```{r load_data, echo=FALSE, message = FALSE, warning = FALSE}

# Load data for plotting, specifying the type for each column
doacs_df <- readr::read_csv(here::here("data/DOACs_data.csv"), 
                            col_types = "Dcccddd")

# CCG_code_lookup (needed for maps)
ccg_lookup <- readr::read_csv(here::here("data/CCG_lookup.csv")) %>% 
  janitor::clean_names() %>% 
  select(ccg21cd, ccg21cdh)

# Load in geoJSON file with ccg shapes (needed for maps)
CCG_shapes_df <- sf::read_sf(here::here("data/CCG_shapes_2021_04.geojson")) %>% 
  janitor::clean_names() %>% 
  select(-ccg21nm)

```


## Histogram

```{r gg_histogram, echo = TRUE}

# Prepare data


# Create chart

nice_hist <- iris %>%
  ggplot() +
  geom_histogram(aes(x = Sepal.Width, fill = Species), 
                 binwidth = 0.2, color = "black") +
  # Add a line on the x axis
  geom_hline(yintercept = 0, linewidth = 1, colour = "#333333") +
  # Remove extra spacing and set y axis limits
  scale_y_continuous(expand = c(0,0), limits = c(0,40)) +
  # Use NICE colours
  scale_fill_manual(values = c("#228096", "#D07B4D", "#EAD054")) +
  # Add nice theme and set labels
  nice_gg_simple() +
  labs(title = "Species setosa has the longest sepal length",
       subtitle = "Distribution of sepal width by species",
       x = "Sepal Width",
       y = "Frequency")

nice_hist

#finalise_plot(example_plot, source_name = "Source: Iris dataset")

```


## Bar chart

```{r gg_bar, echo = TRUE, message = FALSE, warning = FALSE}

# Prepare data

bar_df <- doacs_df %>% 
  filter(between(date, as.Date("2021-01-01"), as.Date("2021-12-31"))) %>% 
  group_by(chemical) %>% 
  summarise(items = sum(items)/1000000) %>% 
  ungroup()

# Create chart

bar_chart <- bar_df %>%
  ggplot(aes(y = items, x = reorder(chemical, items))) +
  geom_hline(yintercept = 0, linewidth = 1, colour="#333333") +
  scale_y_continuous(expand = c(0,0), limits = c(0,8)) +
  scale_x_discrete(labels = function(x) {stringr::str_wrap(x, width = 10)}) +
  geom_col(fill = "#228096", colour = "#000000") +
  nice_gg_simple(base_size = 12) +
  theme(axis.title.x = ggplot2::element_blank()) +
  labs(title = "Apixaban was the most prescribed DOAC in 2021",
       subtitle = "Total DOAC medicines dispensed in primary care in England, 2020",
       y = "Dispensed items (millions)",
       alt = "example alt text")

bar_chart

```

## Line chart

```{r gg_line, echo= TRUE, message = FALSE, warning = FALSE}

# Prepare data
line_df <- doacs_df %>%
  filter(chemical == "Edoxaban") %>% 
  group_by(date, chemical) %>%
  summarise(items = sum(items)) %>%
  ungroup()

# Create chart
line_chart <- line_df %>%
  ggplot(aes(x = date, y = items)) +
  geom_line(linewidth = 0.8, colour = "#228096") +
  geom_point(size = 2, colour = "#228096") +
  scale_y_continuous(expand = c(0,0), limits = c(0,200000), label = scales::comma) +
  scale_x_date(date_labels = "%b\n%Y", date_breaks = "3 months") +
  geom_hline(yintercept = 0, linewidth = 1, colour="#333333") +
  nice_gg_simple() +
  theme(axis.title.x = ggplot2::element_blank()) +
  labs(title = "Edoxaban prescribing has increased",
       subtitle = "Edoxaban prescribing in England, 2020-2022",
       y = "Dispensed items")

line_chart

```

## Scatter plot

```{r gg_scatter, echo= TRUE, message = FALSE, warning = FALSE}

# Prepare data
scatter_df <- doacs_df %>%
  filter(chemical == "Apixaban",
         date == "2021-07-01") %>% 
  mutate(total_list_size = total_list_size/1000)

# Create chart

scatter_chart <- scatter_df %>%
  ggplot() +
  geom_point(aes(x = total_list_size, y = items), 
             colour = "#228096", size = 2) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05)), 
                     limits = c(0,3000), 
                     label = scales::comma) +
  scale_y_continuous(expand = c(0,0), 
                     limits = c(0,25000), 
                     label = scales::comma) +
  nice_gg_simple() +
  theme(panel.border = ggplot2::element_rect(color = "#000000",
                                               linewidth = 0.3,
                                               fill = NA),
        panel.grid.major.x = ggplot2::element_line(color = "#BFBFBF")) +
  labs(title = "Apixaban prescribing",
       subtitle = "Items of apixaban dispensed in primary care compared to CCG list size, 
July 2021",
       x = "Primary care list size (x1000)",
       y = "Dispensed items")

scatter_chart

```

## Faceted chart

```{r gg_facet, echo= TRUE, message = FALSE, warning = FALSE}

# Prepare data
facet_df <- doacs_df %>%
  filter(chemical %in% c("Apixaban", "Edoxaban", "Warfarin sodium")) %>% 
  group_by(date, chemical) %>%
  summarise(items = sum(items)/1000) %>%
  ungroup()

# Create chart

facet_chart <- facet_df %>%
  ggplot(aes(x = date, y = items,
             color = chemical)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(expand = c(0,0), limits = c(0,800)) +
  scale_x_date(date_labels = "%b\n%Y", date_breaks = "6 months") +
  scale_color_manual(values = c("#228096", "#D07B4D", "#37906D")) +
  facet_wrap(~chemical) +
  nice_gg_simple() +
  theme(panel.border = ggplot2::element_rect(color = "#000000",
                                               linewidth = 0.3,
                                               fill = NA),
        axis.title.x = ggplot2::element_blank(),
        legend.position = "none") +
  labs(title = "This is what a title would look like",
       subtitle = "This is what a subtitle would look like",
       x = "Sepal Length",
       y = "Dispensed items")

facet_chart

```

## Choropleth chart (heatmap)

```{r gg_choropleth, echo= TRUE, message = FALSE, warning = FALSE}

# Prepare data

map_df <- doacs_df %>% 
  filter(between(date, as.Date("2021-01-01"), as.Date("2021-12-31")),
         chemical == "Apixaban") %>% 
  group_by(chemical, ccg_name, ccg_ods) %>% 
  summarise(items_per_1000 = sum(items)/(sum(total_list_size)/1000),
            items = sum(items)) %>% 
  ungroup() %>% 
  left_join(ccg_lookup, by = c("ccg_ods" = "ccg21cdh"))


map_df <- left_join(CCG_shapes_df, map_df, by = c("ccg21cd" = "ccg21cd"))

# Create chart
choropleth_chart <- map_df %>% 
  ggplot() +
  geom_sf(aes(fill = items_per_1000), color = "black", size = 0.3) +
  scale_fill_binned(name = "Items per 1000",
                    low = "#e9e9e9",
                    high = "#004650",
                    na.value = "grey50") +
  labs(title = "Apixaban prescribing in England, 2021",
       subtitle = "Plotted by CCG") +
  theme_void()

choropleth_chart <- map_df %>% 
  mutate(items_per_1000_bin = cut(items_per_1000,
                          breaks = c(0,5,10,15,20,Inf),
                          labels = c("<5","5-10", "10-15", "15-20", "20+"))) %>% 
  ggplot() +
  geom_sf(aes(fill = items_per_1000_bin), color = "black", size = 0.2) +
  scale_fill_viridis_d(guide = guide_legend(
                       title = "Items per 1000",
                       direction = "vertical",
                       title.position = "top",
                       title.hjust = 0.3,
                       #title.vjust = 0.,
                       label.position = "left",
                       keywidth = 1.5,
                       keyheight = 1,
                       reverse = TRUE)) +
  labs(title = "Apixaban prescribing in England, 2021" ,
       subtitle = "Plotted by CCG") +
  theme_void() +
  theme(legend.position = c(0.15, 0.5))

choropleth_chart

```


# Interactive charts

## Choropleth map (Leaflet)

```{r leaflet, echo= TRUE, message = FALSE, warning = FALSE}

# Prepare data
map_df <- doacs_df %>% 
  filter(between(date, as.Date("2021-01-01"), as.Date("2021-12-31")),
         chemical == "Warfarin sodium") %>% 
  group_by(chemical, ccg_name, ccg_ods) %>% 
  summarise(items_per_1000 = sum(items)/(sum(total_list_size)/1000),
            items = sum(items)) %>% 
  ungroup() %>% 
  left_join(ccg_lookup, by = c("ccg_ods" = "ccg21cdh"))


map_df <- left_join(CCG_shapes_df, map_df, by = c("ccg21cd" = "ccg21cd"))

# Create chart

pal <- colorBin("plasma",
                domain = map_df$items_per_1000,
                bins = 5,
                na.color = "#808080")

leaflet_map <- map_df %>% 
    # Set up width and degree of zoom per click
    leaflet(width = 800, 
            options = leafletOptions(zoomDelta = 0.25,
                                     zoomSnap = 0.25)) %>% 
    # Set view to centre on England
    setView(lat = 53,
            lng = -1.5,
            zoom = 6) %>%
    # add basic base tile
    addProviderTiles(provider = "CartoDB.Positron") %>%
    # Add shapes
    addPolygons(fillColor = ~pal(items_per_1000),
                fillOpacity = 0.3,
                color = "#393939",
                weight = 1.5,
                opacity = 1,
                layerId = ~ccg_name,
                # Add functionality to highlight shape when hovered over
                highlight = highlightOptions(weight = 3,
                                             color = "#222222",
                                             fillOpacity = 0.7,
                                             bringToFront = TRUE),
                # Add labels upon hover
                label = ~lapply(paste0("<strong>Name: </strong>", ccg_name,
                                       "<br><strong>Items per 1000 population: </strong> ", 
                                       round(items_per_1000, 2)),
                                htmltools::HTML),
                labelOptions = labelOptions(textsize = "12px",
                                            style = list("font-family" = "Arial"))) %>%
    clearControls() %>%
    addLegend(position = "bottomleft",
              pal = pal,
              values = ~items_per_1000,
              title = "Items per 1000 population",
              opacity = 0.7)

leaflet_map

```
