---
title: "NICE visual cookbook for R graphics"
author: "Impact team"
date: "Last updated: November 2022"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dev = "ragg_png")
here::i_am("reports/NICE_cookbook.Rmd")

```

<br>
# How to create NICE style graphics

The `niceRplots` package and cookbook make the process of creating publication-ready graphics in the NICE style a more reproducible process, as well as making it easier for people new to R to create graphics. 

Within this cookbook, we will demonstrate how to use the functions included within the `niceRplots` package. We will also provide examples of plots that have been created in the NICE style. 

<br>
## Load the relevant packages

Before we can start making charts, we first need to load all of the relevant packages. We wont load every package here, just the ones that we will use the most often. It is good practice to limit the number of loaded packages to prevent masking conflicts. For this reason, we will often use double colons `::` to access specific functions within a package (e.g. `readr::read_csv()`), rather than loading the full package using `library(pkg)`.

```{r pkg_load, message = FALSE}

# Load in the relevant packages
library(dplyr)
library(ggplot2)
library(leaflet)

devtools::load_all()

# Set up the NICE colour palette
nice_colours <- c("#228096",
                  "#D07B4D",
                  "#EAD054",
                  "#00436C",
                  "#37906D")

```
<br>

## How to style your charts

As an example we will create a simple bar chart using the `iris` dataset. We will first create a basic chart, and then add some additional formatting. We can see the two charts below with and without formatting.

```{r format_chart, echo = TRUE, fig.show="hold", out.width="50%"}

# Wrangle data for plotting
iris_bar_df <- iris %>% 
  group_by(Species) %>% 
  summarise(Sepal_Width = mean(Sepal.Width)) %>% 
  ungroup()

# Create basic bar chart (left)
iris_bar <- ggplot(iris_bar_df) + 
  geom_col(aes(x = Species, y = Sepal_Width), ) +
  labs(title = "Basic chart title",
       subtitle = "Basic chart subtitle",
       x = "Species",
       y = "Sepal Width")

# Create formatted chart (right)
iris_bar_formatted <- ggplot(iris_bar_df) + 
  # Adjust bar fill and outline colour
  geom_col(aes(x = Species, y = Sepal_Width), fill = nice_colours[1], colour = "#000000")+
  # Add baseline to x axis
  geom_hline(yintercept = 0, linewidth = 1, colour = "#333333") +
  # set y-axis limits and remove padding
  scale_y_continuous(expand = c(0, 0), limits = c(0, 4)) +
  # Add labels 
  labs(title = "Formatted chart title",
       subtitle = "Formatted chart subtitle ",
       y = "Sepal width")

iris_bar
iris_bar_formatted

```
<br>

We can now apply the NICE theme using the `nice_gg_simple()` function from the `niceRplots` package, specifying the base font for the plot to be of size 12. The `nice_gg_simple()` function alters several aspects of the chart to make it consistent with the [NICE brand guidelines](https://indepth.nice.org.uk/our-brand/index.html). These are as follows:

* **Fonts**: changes the title font to Lora SemiBold, and all other text to Inter Regular. 

* **Text scaling**: applies a text hierarchy to the chart using relative text sizes. Changing the `base_size` argument will change the text size throughout the plot. Titles and subtitles will be automatically scaled relative to this base_size to maintain the text hierarchy. This is useful if you want to change the chart size, as all text can be simply adjusted by changing the `base_size` argument. 

* **Colour scheme**: changes line and text colours to be consistent with the NICE brand colour palette. This is particularly noticeable when creating faceted charts (shown later in cookbook). It should be noted that this will not add colour to the plotted data, this will still need to be added manually when creating the chart. For instance, a fill colour was added in the `geom_col()` argument in the above example. 

* **Margins**: adjusts the margins around the title, subtitle, and axis titles to give everything space to breathe.

* **Background**: removes the panel background and borders.

* **Other**: Applies major grid lines along the y-axis. This will be appropriate for most charts, but may need to be disabled in others by applying `theme(panel.grid.major.y = ggplot2::element_blank())`


The `nice_gg_theme` function isn't designed to modify every aspect of the chart, but to apply a general theme that will maintain consistency with the NICE style guide across all charts. Many aspects of the chart will still need to be modified manually. For example, in the below chart we may choose to apply an additional theme argument to remove the x-axis title. 

```{r theme_chart, echo = TRUE}

# Apply the NICE theme
iris_bar_themed <- iris_bar_formatted +
  # Apply nice theme, set base text size to 12
  nice_gg_simple(base_size = 12) +
  # remove x axis title as this is not needed
  theme(axis.title.x = ggplot2::element_blank())

iris_bar_themed

```
<br>

## How to finalise your charts

Once a plot has been created and styled using `nice_gg_simple()`, the next step is to apply the `finalise_plot()` function. This function will create a footer containing information on the data source, as well as adding the NICE logo to the bottom right. It will then left-align the title, subtitle and footer. The `finalise_plot()` function has 3 arguments:

* `plot_name`: the name of the plot object, in the example below we have saved ours as `iris_bar_themed`.

* `source_name`: the source of the underlying data. We have used data from the `iris` dataset.

* `logo`: this will determine whether the NICE logo is included in the footer. It can have a value of "NICE" to include the logo, or "none" to leave it blank.

```{r finalise, echo = TRUE}

# Apply the finalise plot function
finalised_plot <- finalise_plot(iris_bar_themed,
                                source_name = "Iris dataset",
                                logo = "NICE")

finalised_plot
```
<br>

## How to save your finalised plots

After applying the `finalise_plot()` function, the chart can be saved using the `ggsave()` function.

```{r save_plot, echo = TRUE, eval = FALSE}

# Save plot using the ggsave() function 
ggsave("finalised_plot.png", finalised_plot)

```
<br>

# Static charts using ggplot2

The sections below provide examples of how to make different types of static chart using the ggplot2 package. These will demonstrate how to format different charts and apply the NICE theme. For some examples we will continue to use R's built in `iris` dataset. For others we will use data on the monthly Sub-ICB location level prescribing of 5 direct oral anticoagulants (DOACS) in England between February 2017 and January 2022. This was downloaded from [OpenPrescribing.net](https://openprescribing.net/).

```{r load_data, echo = TRUE, message = FALSE, warning = FALSE}

# Load in DOACs data
doacs_df <- readr::read_csv(here::here("data/DOACs_data_full_new.csv"),
                             col_types = "Dcccccdddd")

```
<br>

## Histogram

```{r gg_histogram, echo = TRUE}

# Create chart
nice_hist <- iris %>%
  ggplot() +
  geom_histogram(aes(x = Sepal.Width, fill = Species), 
                 binwidth = 0.2, color = "black") +
  # Add a line on the x axis
  geom_hline(yintercept = 0, linewidth = 1, colour = "#333333") +
  # Remove extra spacing and set y axis limits
  scale_y_continuous(expand = c(0,0), limits = c(0,40)) +
  # Use NICE colours
  scale_fill_manual(values = nice_colours) +
  # Add nice theme and set labels
  nice_gg_simple() +
  labs(title = "Species setosa has the longest sepal length",
       subtitle = "Distribution of sepal width by species",
       x = "Sepal Width",
       y = "Frequency")

nice_hist

#finalise_plot(example_plot, source_name = "Source: Iris dataset")

```
<br>

## Bar chart

```{r gg_bar, echo = TRUE, message = FALSE, warning = FALSE}

# Prepare data, get total items dispensed for each medicine in 2021 (in millions)
bar_df <- doacs_df %>% 
  filter(between(date, as.Date("2021-01-01"), as.Date("2021-12-31"))) %>% 
  group_by(chemical) %>% 
  summarise(items = sum(items)/1000000) %>% 
  ungroup()

# Create chart
bar_chart <- bar_df %>%
  # Plot bars in decreasing order of items
  ggplot(aes(y = items, x = reorder(chemical, -items))) +
  # Add baseline to x axis
  geom_hline(yintercept = 0, linewidth = 1, colour = "#333333") +
  # set y-axis limits and remove padding
  scale_y_continuous(expand = c(0,0), limits = c(0,8)) +
  # set x-axis labels to wrap to prevent overlapping
  scale_x_discrete(labels = function(x) {stringr::str_wrap(x, width = 10)}) +
  # Add bars, add fill and border colours
  geom_col(fill = nice_colours[1], colour = "#000000") +
  # Apply NICE theme
  nice_gg_simple(base_size = 12) +
  # Remove the x-axis title
  theme(axis.title.x = ggplot2::element_blank()) +
  # Add labels
  labs(title = "Apixaban was the most prescribed DOAC in 2021",
       subtitle = "Total DOAC medicines dispensed in primary care in England, 2020",
       y = "Dispensed items (millions)")

bar_chart

```
<br>

## Line chart

```{r gg_line, echo= TRUE, message = FALSE, warning = FALSE}

# Prepare data
line_df <- doacs_df %>%
  filter(chemical == "Edoxaban") %>% 
  group_by(date, chemical) %>%
  summarise(items = sum(items)) %>%
  ungroup()

# Create chart
line_chart <- line_df %>%
  ggplot(aes(x = date, y = items)) +
  # Add baseline to x axis
  geom_hline(yintercept = 0, linewidth = 1, colour="#333333") +
  # Add the vertical dashed line to show lockdown date. Added this before plotting the 
  # data so that it sits on a layer behind the line and points
  geom_vline(xintercept = as.Date("2020-03-01"), linewidth = 0.6, linetype = "dashed") +
  # Add the line and points, make both teal
  geom_line(linewidth = 0.8, colour = nice_colours[1]) +
  geom_point(size = 2, colour = nice_colours[1]) +
  # Set y-axis limits and remove padding, add commas to axis labels
  scale_y_continuous(expand = c(0,0), limits = c(0,200000), label = scales::comma) +
  # Set x-axis labels to show as month year, and to occur at 6-monthly intervals
  scale_x_date(date_labels = "%b\n%Y", date_breaks = "6 months") +
  # Apply NICE theme
  nice_gg_simple() +
  # Remove x-axis title
  theme(axis.title.x = ggplot2::element_blank()) +
  # Add labels
  labs(title = "Edoxaban prescribing has increased since 2017",
       subtitle = "Edoxaban prescribing in England, 2017-2022",
       y = "Dispensed items") +
  # Add annotation for dashed line
  annotate("text", x = as.Date("2020-02-01"), y = 180000, 
           label = "National lockdown\non 23 March", hjust = "right")

line_chart

```
<br>

## Scatter plot

```{r gg_scatter, echo= TRUE, message = FALSE, warning = FALSE}

# Prepare data
scatter_df <- doacs_df %>%
  filter(chemical == "Apixaban",
         date == "2021-07-01") %>% 
  mutate(total_list_size = total_list_size/1000)

# Create chart
scatter_chart <- scatter_df %>%
  ggplot(aes(x = total_list_size, y = items)) +
  # Add the points and make them teal
  geom_point(colour = nice_colours[1], size = 2) +
  # set x-axis limits, adjust padding and add commas to axis labels
  scale_x_continuous(expand = expansion(mult = c(0, 0.05)), 
                     limits = c(0,3000), label = scales::comma) +
  # set y-axis limits and remove padding, add commas to axis labels
  scale_y_continuous(expand = c(0,0), limits = c(0, 25000), label = scales::comma) +
  # Apply NICE theme
  nice_gg_simple() +
  # Add a panel border and vertical grid lines
  theme(panel.border = ggplot2::element_rect(color = "#000000",
                                             linewidth = 0.3,
                                             fill = NA),
        panel.grid.major.x = ggplot2::element_line(color = "#BFBFBF")) +
  # Add labels
  labs(title = "Apixaban prescribing",
       subtitle = "Items of apixaban dispensed in primary care compared to sub-ICB\nlocation list size, July 2021",
       x = "Primary care list size (x1000)",
       y = "Dispensed items")

scatter_chart

```
<br>

## Faceted chart

```{r gg_facet, echo= TRUE, message = FALSE, warning = FALSE}

# Prepare data
facet_df <- doacs_df %>%
  filter(chemical %in% c("Apixaban", "Edoxaban", "Warfarin sodium"),
         between(date, as.Date("2020-01-01"), as.Date("2021-12-31"))) %>% 
  group_by(date, chemical) %>%
  summarise(items = sum(items)/1000) %>%
  ungroup()

# Create chart
facet_chart <- facet_df %>%
  ggplot(aes(x = date, y = items, color = chemical)) +
  # Add lines
  geom_line(linewidth = 1) +
  # Add points
  geom_point(size = 2) +
  ## set y-axis limits and remove padding
  scale_y_continuous(expand = c(0,0), limits = c(0,800)) +
  # set x-axis labels to show as month year, and to occur at 6-monthly intervals
  scale_x_date(date_labels = "%b\n%Y", date_breaks = "6 months") +
  # Manually set the colour of the lines and points
  scale_color_manual(values = nice_colours) +
  # APply facet wrapping to put each chemical into an individual plot
  facet_wrap(~chemical) +
  # Apply the NICE theme
  nice_gg_simple() +
  # Add a panel border, remove x axis title and legend
  theme(panel.border = ggplot2::element_rect(color = "#000000",
                                               linewidth = 0.3,
                                               fill = NA),
        axis.title.x = ggplot2::element_blank(),
        legend.position = "none") +
  # Add labels
  labs(title = "Trends in anti-coagulant prescribing",
       subtitle = "Prescribing of 3 anti-coagulants in England, 2020-2021",
       x = "Sepal Length",
       y = "Dispensed items (thousands)")

facet_chart

```
<br>

## Choropleth chart (heatmap)

```{r gg_choropleth, echo= TRUE, message = FALSE, warning = FALSE}

# Prepare data
map_df <- doacs_df %>% 
  filter(between(date, as.Date("2021-01-01"), as.Date("2021-12-31")),
         chemical == "Apixaban") %>% 
  group_by(chemical, name, ods_code, gss_code) %>% 
  summarise(items_per_1000 = sum(items)/(sum(total_list_size)/1000),
            items = sum(items)) %>% 
  ungroup()

# Load in the dataframe containing the shape files
sloc_shapes_df <- sf::read_sf(here::here("data/sub_icb_shapes.geojson")) %>% 
  janitor::clean_names() %>% 
  select(-objectid)

sf::sf_use_s2(FALSE)
nat_shape <- sf::st_union(sloc_shapes_df)

# Join to dataframe containing shape files. Ensure the dataframe containing the 
# shape files is the first argument in the join, as we want to preserve the class of this table. 
# If not the column containing the shapes will be dropped.
map_df <- left_join(sloc_shapes_df, map_df, by = c("sicbl22cd" = "gss_code"))

# Set up palette
pal <- colorRampPalette((c("#BFBFBF", nice_colours[1])))(5)
#pal <- colorRampPalette((c("#228096", "#BFBFBF","#EAD054")))(5) # 
#pal <- viridis::plasma(5) # viridis palette

# Create chart 
choropleth_chart <- map_df %>% 
  # Set up a column with categorical labels for each bin
  mutate(items_per_1000_bin = cut(items_per_1000,
                                  breaks = c(0,5,10,15,20,Inf),
                                  labels = c("<5","5-10", "10-15", "15-20", "20+"))) %>% 
  ggplot() +
  # Plot the shapes
  geom_sf(aes(fill = items_per_1000_bin), color = "#FFFFFF", lwd = 0.1) +
  # Use the viridis palette and adjust the legend style
  scale_fill_manual(values = pal, 
                    na.value = "#808080",
                    guide = guide_legend(
                      title = "Items per 1000",
                      label.position = "left",
                      reverse = TRUE)) +
  geom_sf(data = nat_shape, fill = NA, color = "black", lwd = 0.1) +
  # Add the NICE theme
  nice_gg_simple() +
  # Postion the legend
  theme(legend.position = c(0.15, 0.5),
        axis.text = element_blank(),
        axis.ticks.x = ggplot2::element_blank(),
        legend.title = ggplot2::element_text(family = "Inter",
                                             size = 12,
                                             color = "#000000"),
        legend.text.align = 1) +
  # Add labels
  labs(title = "Apixaban prescribing in England, 2021",
       subtitle = "Plotted by CCG")

choropleth_chart

```
<br>

# Interactive charts

## Choropleth map (Leaflet)

```{r leaflet, echo= TRUE, message = FALSE, warning = FALSE}

# Using same data as above

# Create chart

pal <- colorBin("plasma",
                domain = map_df$items_per_1000,
                bins = 5,
                na.color = "#808080")

leaflet_map <- map_df %>% 
    # Set up width and degree of zoom per click
    leaflet(width = 800, 
            options = leafletOptions(zoomDelta = 0.25,
                                     zoomSnap = 0.25)) %>% 
    # Set view to centre on England
    setView(lat = 53,
            lng = -1.5,
            zoom = 6) %>%
    # add basic base tile
    addProviderTiles(provider = "CartoDB.Positron") %>%
    # Add shapes
    addPolygons(fillColor = ~pal(items_per_1000),
                fillOpacity = 0.3,
                color = "#393939",
                weight = 1.5,
                opacity = 1,
                layerId = ~ods_code,
                # Add functionality to highlight shape when hovered over
                highlight = highlightOptions(weight = 3,
                                             color = "#222222",
                                             fillOpacity = 0.7,
                                             bringToFront = TRUE),
                # Add labels upon hover
                label = ~lapply(paste0("<strong>Name: </strong>", sicbl22nm,
                                       "<br><strong>Items per 1000 population: </strong> ", 
                                       round(items_per_1000, 2)),
                                htmltools::HTML),
                labelOptions = labelOptions(textsize = "12px",
                                            style = list("font-family" = "Arial"))) %>%
    clearControls() %>%
    addLegend(position = "bottomleft",
              pal = pal,
              values = ~items_per_1000,
              title = "Items per 1000 population",
              opacity = 0.7)

leaflet_map

```
